---
title: "MLM Sample Sizes"
author: "Scott Zeger"
date: "4/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## This note presents one simple approach to calculating the power of a study to detect a non-zero regression coefficient using a 2-level model. The user must specify
*the between cluster variance of X_bar_i (VarX_w)
*the within-cluster variance of X_ij - X_bar_i (VarX_b) 
*note: Var(X_ij) = VarX_w + VarX_b  
*the 2x2 covariance matrix (D) for the random intercept and slope (D_jk = 0 is allowed for a single random effect)
* the variance of the residuals (sigma2)
* the number of clusters (m)
* the mean (mean_n_i) of the Poisson distribution that generates the within-cluster samples sizes (n_i)


```{r data input}
library(Matrix)
sigma2_0 = 1
D0=matrix(c(1,0,0,1),ncol=2)
mean_n0_i = 10
m0=25
# generate n_i values
results=NULL
case=NULL
for (j in 0:2) {
  mean_n_i = mean_n0_i*(2^j)
  n_min=2
  n_i = rpois(m,mean_n_i) + n_min
  N=sum(n_i)
  VarX_w = 1
  VarX_b=3
  # generate vector of IDs
id = c(rep(1:m,n_i))
# generate X-matrix
  X=cbind(rep(1,N),rnorm(N)*sqrt(VarX_w),rep(rnorm(m),n_i)*sqrt(VarX_b))
  Z=X[,1:2]
#V_inv=list()

  for (k in 0:2) {
    D=(2*k)*D0
    result= var_beta(X,Z,id,D,sigma2_0,m)
    results=cbind(results,sqrt(diag(result)))
    case=c(case,(j*10+k))
  }
# results=NULL
# for (j in 0:2) {
#   sigma2=sigma2_0*(2^j)
#   result= var_beta(X,Z,id,D0,sigma2,m)
#   results=cbind(results,sqrt(diag(result)))
# }
# print(results)
#
}
colnames(results)=as.character(case)
rownames(results)=c("beta_0","beta_1w","beta_1b")
print(results)
# 
var_beta=function(X,Z,id,D,sigma2=1,m=10){
#  this function calculates the covariance matrix of regression coefficients for a 2-level model
#
M = matrix(0,nrow=ncol(X),ncol=ncol(X))
for (i in 1:m) {
  Z_i = Z[id==i,]
  X_i = X[id==i,]
  V_i_inv=solve(Z_i%*%D%*%t(Z_i)+diag(rep(sigma2,n_i[i])))
  #V_inv=c(V_inv,V_i_inv)
  M_i = t(X_i)%*%V_i_inv%*%X_i
  M=M+M_i
}
var_beta=solve(M)
return(var_beta=var_beta)
}
```
