---
title: "Lecture 4"
author: "Lacey Etzkorn"
date: "April 2, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(gridExtra)
library(lme4)
df <- read_csv(file = file.path("Data","gcse.csv"))
df <- df %>%
group_by(school) %>%
mutate(schoolcount = n()) %>%
ungroup
```

###EDA

```{r}
# How many total subjects in the dataset
nrow(df)
```

### What is the distribution of number of students in each school

```{r}
df %>% filter(student == 1) %>% with(summary(schoolcount))
```

### What is the distribution of the gcse and lrt scores

```{r}
df %>% select(gcse, lrt) %>% lapply(summary)
```

### What is the relationship between gcse and lrt for few schools
```{r}
df %>% 
filter(school <= 12) %>%
ggplot(aes(lrt, gcse)) +
geom_point() +
stat_smooth(method="lm") +
xlab("LRT Score") +
ylab("GCSE Score") +
facet_wrap("school") +
theme_classic()
```

### Create SLR of gcse on lrt for each school and plot the fitted lines

```{r}
df %>% 
ggplot(aes(lrt, gcse)) +
geom_point(color = "grey80") +
stat_smooth(aes(group = school), 
	method="lm", se=F, size = 0.5) +
xlab("LRT Score") +
ylab("GCSE Score") +
theme_classic()
```

### Create the same graph we saw in Lecture 3 (between and within effects on same graph)

```{r}
df <- df %>%
group_by(school) %>%
mutate(mean_lrt = mean(lrt),
       lrt_within = lrt - mean_lrt, 
       mean_gcse = mean(gcse),
       gcse_within = gcse - mean_gcse) %>%
ungroup() %>%
mutate(lrt_scaled = lrt_within + mean(lrt),
       gcse_scaled = gcse_within + mean(gcse))
ggplot()+
geom_point(data = df, aes(x = lrt, y = gcse), color = "grey80") +
geom_point(data = filter(df, student == 1),
           aes(x = mean_lrt, y = mean_gcse), color = "red") +
stat_smooth(data = df, aes(x = lrt_scaled, y = gcse_scaled), 
	method = "lm", se = F, color = "blue") +
stat_smooth(data = filter(df, student == 1),
            aes(x = mean_lrt, y = mean_gcse), 
	method = "lm", color = "darkred", se = F) +
stat_smooth(data = df,
            aes(x = lrt, y = gcse), 
	method = "lm", color = "black", se = F) +
xlab("London Reading Test Score") + 
ylab("GCSE Score") +
theme_classic() +
# add our own legend
annotate("point", x = rep(80,5), y = c(55, 50, 45,40,35), 
         color = c("black", "blue", "darkred", "red","grey80"),
         shape = c(45,45,45,19,19), size = c(10,10,10,2,2)) +
annotate("text", x = rep(80,5)+1, y = c(55, 50, 45,40,35), 
         label = c("Total Effect", "Within Effect", "Between Effect",
                   "School Averages","Observed Scores"),
         hjust = 0)
```

### Possible models

```{r}
mod1 <- lmer(data = df, gcse~lrt + (1|school))
mod2 <- lmer(data = df, gcse~lrt+mean_lrt + (1|school))
mod3 <- lmer(data = df, gcse~lrt_within+mean_lrt + (1|school))
mod4 <- lmer(data = df, gcse~lrt_within + (1|school))
mod5 <- lmer(data = df, gcse~mean_lrt + (1|school))
list(mod1, mod2, mod3, mod4, mod5) %>%
lapply(fixef)
```

### Interpretation of Models 4 and 5

```{r}
df %>% select(lrt,lrt_within,mean_lrt,girl,schgend) %>% cor()
```

### Demonstrate the appropriate adjustment if you don't center or grand mean center

```{r}
df <- df %>%
mutate(lrt_grandmean = lrt - mean(lrt),
       girl_grandmean = girl - mean(girl)) %>%
group_by(school) %>%
mutate(mean_girl = mean(girl),
       girl_within = girl - mean_girl)
mod1 <- lmer(data = df, gcse~factor(schgend) + (1|school))
mod2 <- lmer(data = df, gcse~factor(schgend)+lrt+girl + (1|school))
mod3 <- lmer(data = df, gcse~factor(schgend)+lrt_grandmean + girl_grandmean + (1|school))
mod3 <- lmer(data = df, gcse~factor(schgend)+lrt_within + girl_within + (1|school))
list(mod1, mod2, mod3, mod4) %>%
lapply(fixef)
```
