---
title: "Lecture 3"
author: "Lacey Etzkorn"
date: "April 2, 2019"
output: html_document
---

As an R programmer, it might be useful for you to learn to use R markdown. We are going to begin writing lecture notes using .rmd files. 

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(gridExtra)
df <- read_csv(file = file.path("Data","gcse.csv"))
head(df)
```

### EDA:

##### 1)  Quantify the relationship between GCSE score and LRT score.

```{r}
ggplot(df,aes(x = lrt, y = gcse))+
geom_point() +
stat_smooth(method = "loess") +
xlab("London Reading Test Score") + 
ylab("GCSE Score") +
theme_classic()
```

##### 2) Within a school, quantify the relationship between GCSE score and LRT score

##### 3) Does the context of the school matter?  I.e. do students from schools with higher school-average LRT scores fair better than similar students in schools with lower school-average LRT scores

Start by decomposing the information in LRT into within and between. Create a the mean gcse across schools, to quantify the between effect

```{r}
df <- df %>%
group_by(school) %>%
mutate(mean_lrt = mean(lrt),
       lrt_within = lrt - mean_lrt, 
       mean_gcse = mean(gcse),
       gcse_within = gcse - mean_gcse) %>%
ungroup() %>%
mutate(lrt_scaled = lrt_within + mean(lrt),
       gcse_scaled = gcse_within + mean(gcse))
head(df)
```

Create two graphs:  gcse_within vs. lrt_within and mean_gcse vs. mean_lrt

```{r, fig.width = 8}
within.plot <- 
ggplot(df, aes(y=gcse_within, x=lrt_within)) +
geom_point() + 
stat_smooth(method = "loess") + 
xlab("LRT score - School-mean LRT score") +
ylab("GCSE score - School-mean GCSE score") + 
ggtitle("Within School Effect of LRT on GCSE")+
theme_classic()

between.plot <- 
df %>%
# we only need one student per school to make this plot
filter(student == 1) %>% 
ggplot(aes(y=mean_gcse, x=mean_lrt)) +
geom_point() + 
stat_smooth(method = "loess") + 
xlab("School-mean LRT score") +
ylab("School-mean GCSE score") + 
ggtitle("Between School Effect of LRT on GCSE") + 
theme_classic()

grid.arrange(within.plot, between.plot, nrow = 1)
```

Overlay the between and within group effects on the same graph

```{r}
ggplot()+
geom_point(data = df, aes(x = lrt, y = gcse), color = "grey80") +
geom_point(data = filter(df, student == 1),
           aes(x = mean_lrt, y = mean_gcse), color = "red") +
stat_smooth(data = df, aes(x = lrt_scaled, y = gcse_scaled), 
	method = "lm", se = F, color = "blue") +
stat_smooth(data = filter(df, student == 1),
            aes(x = mean_lrt, y = mean_gcse), 
	method = "lm", color = "darkred", se = F) +
stat_smooth(data = df,
            aes(x = lrt, y = gcse), 
	method = "lm", color = "black", se = F) +
xlab("London Reading Test Score") + 
ylab("GCSE Score") +
theme_classic() +
# add our own legend
annotate("point", x = rep(80,5), y = c(55, 50, 45,40,35), 
         color = c("black", "blue", "darkred", "red","grey80"),
         shape = c(45,45,45,19,19), size = c(10,10,10,2,2)) +
annotate("text", x = rep(80,5)+1, y = c(55, 50, 45,40,35), 
         label = c("Total Effect", "Within Effect", "Between Effect",
                   "School Averages","Observed Scores"),
         hjust = 0)
```

##### Create a dataset with no ecological fallacy

```{r}
set.seed(88491)
mod1 <- lm(data = df, gcse~lrt_within+mean_lrt)
df <- df %>%
mutate(error = rnorm(nrow(df), 0, sigma(mod1)),
       new_gcse = error + mod1$coef[1] + mod1$coef[2]*lrt_within + mod1$coef[3]*mean_lrt) %>%
group_by(school) %>%
mutate(mean_gcse2 = mean(new_gcse))

ggplot()+
geom_point(data = df, aes(x = lrt, y = new_gcse), color = "grey80") +
geom_point(data = filter(df, student == 1),
           aes(x = mean_lrt, y = mean_gcse), color = "red") +
stat_smooth(data = df, aes(x = lrt_scaled, y = new_gcse), 
	method = "lm", se = F, color = "blue") +
stat_smooth(data = filter(df, student == 1),
            aes(x = mean_lrt, y = mean_gcse), 
	method = "lm", color = "darkred", se = F) +
stat_smooth(data = df,
            aes(x = lrt, y = new_gcse), 
	method = "lm", color = "black", se = F) +
xlab("London Reading Test Score") + 
ylab("GCSE Score") +
theme_classic() +
# add our own legend
annotate("point", x = rep(80,5), y = c(55, 50, 45,40,35), 
         color = c("black", "blue", "darkred", "red","grey80"),
         shape = c(45,45,45,19,19), size = c(10,10,10,2,2)) +
annotate("text", x = rep(80,5)+1, y = c(55, 50, 45,40,35), 
         label = c("Total Effect", "Within Effect", "Between Effect",
                   "School Averages","Simulated Scores"),
         hjust = 0)
```

Replicate with between effect is 3 times the within effect just to demonstrate.

```{r}
set.seed(77432)
mod1 <- lm(data = df, gcse~lrt_within+mean_lrt)
df <- df %>%
ungroup() %>%
mutate(error = rnorm(nrow(df), 0, sigma(mod1)),
       new_gcse = error + mod1$coef[1] + mod1$coef[3]/3*lrt_within + mod1$coef[3]*mean_lrt) %>%
group_by(school) %>%
mutate(mean_gcse2 = mean(new_gcse))

ggplot()+
geom_point(data = df, aes(x = lrt, y = new_gcse), color = "grey80") +
geom_point(data = filter(df, student == 1),
           aes(x = mean_lrt, y = mean_gcse), color = "red") +
stat_smooth(data = df, aes(x = lrt_scaled, y = new_gcse), 
	method = "lm", se = F, color = "blue") +
stat_smooth(data = filter(df, student == 1),
            aes(x = mean_lrt, y = mean_gcse), 
	method = "lm", color = "darkred", se = F) +
stat_smooth(data = df,
            aes(x = lrt, y = new_gcse), 
	method = "lm", color = "black", se = F) +
xlab("London Reading Test Score") + 
ylab("GCSE Score") +
theme_classic() +
# add our own legend
annotate("point", x = rep(80,5), y = c(55, 50, 45,40,35), 
         color = c("black", "blue", "darkred", "red","grey80"),
         shape = c(45,45,45,19,19), size = c(10,10,10,2,2)) +
annotate("text", x = rep(80,5)+1, y = c(55, 50, 45,40,35), 
         label = c("Total Effect", "Within Effect", "Between Effect",
                   "School Averages","Simulated Scores"),
         hjust = 0)
```

##### 4) Are there differences in performance across school type (mixed gender, girls only and boys only)?

```{r}
df$gender <- factor(df$schgend, labels = c("Mixed", "Boys", "Girls"))
df %>%
filter(student==1) %>%
ggplot() +
geom_boxplot(aes(x = gender, fill = gender, y = mean_gcse)) +
theme_classic()
```


##### 5) Within a school, does student gender modify the relationship between GCSE score and LRT score?
```{r}
ggplot(df,aes(y = gcse_within, x = lrt_within)) +
geom_point() +
stat_smooth(method = "loess") +
facet_wrap("girl") +
xlab("LRT score - School-mean LRT score") +
ylab("GCSE score - School-mean GCSE score") +
theme_classic()
```


##### 6) Does school type modify the relationship between GCSE score and LRT score?

```{r}
ggplot(df,aes(y = gcse_within, x = lrt_within)) +
geom_point() +
stat_smooth(method = "loess") +
facet_wrap("gender") +
xlab("LRT score - School-mean LRT score") +
ylab("GCSE score - School-mean GCSE score") +
theme_classic()
```


```{r}
ggplot(df,aes(y = gcse_within, x = lrt_within)) +
geom_point(color = "grey80") +
stat_smooth(method = "loess", aes(color = gender), se=FALSE) +
xlab("LRT score - School-mean LRT score") +
ylab("GCSE score - School-mean GCSE score") +
theme_classic()
```
